Using Google Community Mobility Reports to investigate the growth of COVID-19 in the United States

### 2020-05-11 Covid-19-Google-CMR-US/Response-to-Reviewers-v2.Rmd

---
title: "Response to Reviewers"
author: "Antonio Paez"
date: "5/11/2020"
output: pdf_document
---

I would like to thank two anonymous reviewers for their comments and suggestions for revisions. In this letter I respond to those comments and explain how the paper was revised in response.

\newpage

# Reviewer 1

This is a well-written and timely paper. I only have some comments regarding the interpretation. So Table 3 indicates that park-related mobility (I assume this refers to walking and cycling in a park? this could be specified) results in a new/more cases of Covid 19, while residential mobility (mobility in own neighborhood?) results in a lower amount of Covid 19 cases? I think this could be stated somewhat more clear/ easier to understand in the paper.

>*Thank you for your positive assessment of the paper, and comments for improving it. The parks mobility index refers to mobility trends for places like local parks, national parks, public beaches, marinas, dog parks, plazas, and public gardens. Following a recommendation from Reviewer 2 I have replaced "residential mobility" with "work-related mobility", which is easier to interpret. Hopefully in the revision the results are more clearly stated.*

I would also suggest adding a sentence (before of after table 3) like: ‘These results indicate that traveling outside the residential neighborhood increases the chance of getting infected by COVID-19, and therefore aligns with existing social distancing rules’ (or a similar sentence refering to policy implication)

>*Thank you. I added a sentence regarding the policy implications of the analysis.*

\newpage

# Reviewer 2

The question the paper deals with is timely and interesting. However, there are a number of issues. First, the intent of the stay-at-home orders is to reduce transmission from what it would have been otherwise. Cases are going up regradless, but if the policy works, there would be fewer cases. Without a counterfactual of what the number of infections would have been, it is impossible to find meaning from the model. It is the reductions that should be a response variable to mobility changes.

>*That is correct. In lieu of a counterfactual, what we have is differences in compliance with the policy in different states, which are reflected statistically in different rates in the growth of cases. See for example the table below. On March 21, New Jersey had similary mobility with respect to parks and work as Idaho. However, these two states had a different starting point for the trajectory of the pandemic - which is reflected in the number of new cases on that date (both actual and estimated by the model). By April 20, New Jersey had substantially reduced park-related mobility, whereas Idaho's was even higher. The estimated and actual number of new cases grew in the intervening period; however, New Jersey's growth in new cases (from March 21 to April 20) was only 894%, whereas Idaho's was 1030%.*

State | Date | Parks | Work | Estimated New Cases | Actual New Cases |
-| - | - | - | - | - |
New Jersey | March 21 | 1.45 | 0.845 | 441.42 | 440 |
New Jersey | April 20 | 0.728 | 0.413 | 3498.19 | 4377 |
Idaho | March 21 | 1.48 | 0.939 | 23.10 | 23
Idaho | April 20 | 1.63 | 0.588 | 259.82 | 260

Second, the chosen mobility indicators (parks and residence) both increased from baseline in the period of analysis for most states (median > 1 for both variables). Neither captures the policy intent. Third, the residence mobility variable is not well defined as the author indicates in their conclusion. Other variables such as work, transit better capture the policy intent and are better defined.

>*I am grateful for these comments, which made me rethink the selection of variables. Here, it is important to make a distinction between the _intent_ of the policy, and _compliance_ with the policy. While the policy's intent has been to reduce out-of-home activities to limit the contact rate for the virus, compliance has been haphazard, with some states (such as New Jersey in the example above) recording sharper declines in mobility, and others (such as Idaho) where compliance has been more lax.*

>*The above said, I was encouraged by this comment to replace residential mobility (which I struggled with the interpretation of in any case) and use instead work-related mobility, while retaining park-related mobility. Why these two variables? First, as seen in Table 2, the correlations between the variables are very high, which means that after choosing any one, a second mobility index does not bring a lot of new information. The exception is parks, the only variable with R < 0.8 with any of the other mobility indicators. So partly, I chose to keep parks in the analysis for statistical reasons. Secondly, work and parks capture two dimensions of mobility, mandatory and discretionary travel, so conceptually they make sense too. And finally, these two mobility indicators are better defined than residential-based mobility.*

>*The results indicate that the number of new cases generally increases as a function of mobility, but the effect is not linear, given the interactions between work-related and parks-related mobility.*

Finally, even though the author states that “higher levels of mobility tend to be associated with higher number of cases”, that is not what the estimated parameters show within their range in the data for both residence and parks. For example, taking the quadratic function for residences, within the range of 0.34-1.16, infections decline as mobility increases - the opposite of what the policy is supposed to do. It seems that the declines in the included figures are driven by the date variable and interactions with it.

>*There is indeed a strong temporal trend, which is captured by the date and interaction of date with the mobility indicators. The coefficients of the model indicate that the number of new cases increases with park-related mobility, and also increases with work related mobility. The interaction term is negative, which captures the trade-off between these two forms of mobility. As seen in the figures, the trend is stronger in the dimension of parks-related mobility: as trade-offs go, gains due to lower work-related mobility can be more than offset by parks-related mobility.*

### 2020-05-11 Covid-19-Google-CMR-US/Covid-19-Google-CMR-US.Rmd
Updated link to repository
Second version of paper after first round of anonymous reviews.

---
title: Using Google Community Mobility Reports to investigate the growth of COVID-19 in the United States
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
address:
  - code: McMaster University
    address: School of Geography and Earth Sciences, McMaster University, Hamilton, ON, L8S 4K1, Canada
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
   In 2020 Google released a set of Community Mobility Reports (GCMR). These reports are based on the company's location-tracking capabilities and measure changes in mobility with respect to a baseline. This novel source of data offers an opportunity to investigate the way mobility potentially correlates with transmission of COVID-19. Using data from the New York Times on COVID-19 cases and GCMR, this paper presents an analysis of mobility levels and new daily reported cases of COVID-19 by state in the US. The results provide insights about the utility and interpretability of GCMR for COVID-19 research and decision-making.

journal: "Transportation Findings"
date: "`r Sys.Date()`"
bibliography: References.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---

@@

```{r add-day-variable, include=FALSE}
#add day variable:
c19 <- c19 %>%
  mutate(day = yday(date) - min(yday(date)))
```

@@

```{r prepare-variables-for-analysis, include=FALSE}
#replace mobility indicators by their lagged moving averages
c19 <- c19 %>% 
  mutate(retail = Mean_r_and_r_lag11,
         groceries = Mean_g_and_p_lag11,
         parks = Mean_parks_lag11,
         parks2 = Mean_parks_lag11^2,
         transit = Mean_transit_lag11,
         work = Mean_work_lag11,
         residential = Mean_residential_lag11,
         residential2 = Mean_residential_lag11^2,
         parksXresidential = parks * residential,
         day2 = day^2,
         parksXday = parks * day,
         parksXday2 = parks * day2,
         residentialXday = residential * day,
         residentialXday2 = residential * day2,
         NY = ifelse(state == "New York", 1, 0),
         NYxday = NY * day)
```

@@

The main policy tool to control the spread of the COVID-19 pandemic has been stay-at-home orders. Concurrently, numerous efforts exist to track the progress and the impact of the pandemic, and new sources of data include the recently-released Google Community Mobility Reports (GCMR)\footnote{https://www.google.com/covid19/mobility/}, as well as The New York Times repository of COVID-19 data\footnote{https://github.com/nytimes/covid-19-data}. These two open data sets offer novel opportunities to investigate in quasi-real time the relationship between mobility patterns and transmission of COVID-19.

@@

The source document for this paper is an `R` markdown document available from the following repository:

https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US

@@

GCMR use aggregated and anonymized data to chart changes in mobility with respect to different classes of places, namely retail and recreation, groceries and pharmacies, parks, transit stations, workplaces, and residential. Mobility indicators for each of these places are reported as percentage changes from a baseline level, which corresponds to the median value of mobility of identical days of the week during the period between January 3 and Feb 6, 2020. Covid-19 data is compiled by The New York Times based on reports from state and local health agencies.

For analysis, all mobility indicators are centered so that the value of 1 is the base mobility, and a 0.01 deviation corresponds to a 1% change. The incubation time of the disease has been estimated to be between 2 and 12 days (95% interval) by Lauer et al. [-@Lauer2020incubation]. Given this, it is to be expected that any changes in mobility will have a lagged effect on the discovery of new cases. For this reason, lagged moving averages of the mobility indicators are calculated. Furthermore, it is possible that mobility and reports of new cases of COVID-19 are endogenous, if the public adjust their mobility according to reports of the incidence. Therefore, in addition to being consistent with an incubation period, use of lagged indicators also helps to break this potential endogeneity.

```{r descriptive-statistics, echo=FALSE}
data.frame(Variable = c("New Cases",
                        "date",
                        "retail",
                        "groceries",
                        "parks",
                        "transit",
                        "work",
                        "residential"),
           Definition = c("New Daily Cases of COVID-19",
                        "Date",
                        "Retail and Recreation",
                        "Groceries and Pharmacies",
                        "Parks",
                        "Transit Stations",
                        "Workplaces",
                        "Residential"),
           min = c(as.character(min(c19$new_cases)),
                   as.character(min(c19$date)),
                   as.character(round(min(c19$retail), 2)),
                   as.character(round(min(c19$groceries), 2)),
                   as.character(round(min(c19$parks), 2)),
                   as.character(round(min(c19$transit), 2)),
                   as.character(round(min(c19$work), 2)),
                   as.character(round(min(c19$residential), 2))),
           median = c(as.character(median(c19$new_cases)),
                   as.character(median(c19$date)),
                   as.character(round(median(c19$retail), 2)),
                   as.character(round(median(c19$groceries), 2)),
                   as.character(round(median(c19$parks), 2)),
                   as.character(round(median(c19$transit), 2)),
                   as.character(round(median(c19$work), 2)),
                   as.character(round(median(c19$residential), 2))),
           max = c(as.character(max(c19$new_cases)),
                   as.character(max(c19$date)),
                   as.character(round(max(c19$retail), 2)),
                   as.character(round(max(c19$groceries), 2)),
                   as.character(round(max(c19$parks), 2)),
                   as.character(round(max(c19$transit), 2)),
                   as.character(round(max(c19$work), 2)),
                   as.character(round(max(c19$residential), 2))),
           sd = c(as.character(round(sd(c19$new_cases), 2)),
                   as.character(""),
                   as.character(round(sd(c19$retail), 2)),
                   as.character(round(sd(c19$groceries), 2)),
                   as.character(round(sd(c19$parks), 2)),
                   as.character(round(sd(c19$transit), 2)),
                   as.character(round(sd(c19$work), 2)),
                   as.character(round(sd(c19$residential), 2)))) %>%
  kable("latex",
        booktabs = TRUE,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics of the data set") %>%
  kable_styling(latex_options = c("striped", "scale_down", "hold_position")) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

The lagged indicators are calculated as the mean of the mobility indicator using the values from date-minus-12-days to date-minus-2-days. Furthermore, using the cumulative number of reported COVID-19 cases, the total number of new daily cases is calculated. This variable (log-transformed after adding a small constant) is paired to the corresponding lagged moving average of the mobility indicators. The log-transformation is useful to avoid negative values of daily new cases when making predictions. Analysis is based on correlation analysis, multivariate regression, and data visualization. Table \ref{tab:descriptive-statistics} shows the descriptive statistics of the data set. Temporal coverage is from February 27 to April 26, 2020. The maximum number of new daily cases during this period is 12,126.

@@

Table \ref{tab:correlation-analysis} shows that the mobility indicators are highly correlated with each other. This is not surprising: it is well-known that mobility involves time-use trade-offs: the more there is of one kind of mobility, the less time there is available for any other. That said, the strongest correlation with the outcome of interest is for residential-based mobility. To avoid multicollinearity, parks-related mobility is selected as a covariate since it has the lowest correlation with residential-based mobility.

```{r check-correlations, echo=FALSE}
#Check the correlation between the mobility indicators and incidence:
c19 %>% 
  select(log_new_cases, retail, groceries, parks, transit, work, residential) %>%
  cor(use = "pairwise.complete.obs") %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators")  %>%
  kable_styling(latex_options = c("striped", "scale_down", "hold_position")) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

A regression model uses the log of new daily cases as the dependent variable. The covariates are parks- and residential-related mobility, which enter the regression in the form of a second order polynomial expansion. In addition, the date is introduced to account for the temporal trend of the pandemic. Finally, an indicator variable for the state of New York and an interaction with the date are used to distinguish the unusually high incidence of the disease there. The results appear in Table \ref{tab:model-results}. The model provides a good fit to the data and all variables reported are significant at $p<0.05$ or better.

There is an overall temporal trend that indicates a growing number of cases over time, but at a decreasing rate (see negative sign of date^2). New York has on average more new daily cases than the rest of the states, but this has also declined over time (see negative sign of NY x date). Visualization is the most effective way to understand the trend according to the mobility indicators and date. Figure \ref{fig:prediction-plots} shows the prediction surfaces on three different dates: March 21, when the first states began implementing stay-at-home orders; April 5, fifteen days later; and April 20, fifteen more days later, and at a time when some states are considering relaxing stay-at-home orders. On March 21 there were still only minor departures from baseline mobility (recall that these are temporally lagged); the prediction surface at this point is relatively flat. This changes by April 5, when large increases in residential-based mobility are seen, along with greater variations in parks-based mobility. The prediction surface indicates an expectation of a greater number of new daily cases as both classes of mobility increase. On the last date examined, April 20, the trend becomes more steep for park-based mobility, even as this indicator continues to display large variations from the baseline in both directions. In general, higher levels of mobility tend to be associates with a higher number of new daily cases.

```{r model-estimation, include=FALSE}
#This is the regression model (forcing the regression through the intercept):
mod1 <- lm(log_new_cases ~  0 +
             day +
             day2 +
             parks2 + 
             parks + 
             parksXresidential +
             residential +
             residential2 +
             parksXday + 
             #parksXday2 +
             residentialXday +
             residentialXday2 +
             NY + 
             NYxday, 
           data = c19 #%>% filter(date >= "2020-04-01")
           )
summ(mod1)
```

```{r model-results, echo = FALSE}
data.frame(Variable = c("date",
                        "date^2",
                        "parks^2",
                        "parks",
                        "parks x residential",
                        "residential",
                        "residential^2",
                        "parks x date",
                        "residential x date",
                        "residential x date^2",
                        "NY",
                        "NY x date"),
           Coefficient = as.numeric(coefficients(mod1)),
           pval = as.numeric(summary(mod1)$coefficients[,4])) %>%
  mutate(pval = ifelse(pval >= 0.0001, round(pval, 4), "<0.0001")) %>%
  kable("latex",
        booktabs = TRUE,
        digits = 4,
        caption = "\\label{tab:model-results}Results of estimating regression model. Dependent variable is log(New Cases + 0.0001).",
        align = c("l", "c", "c"),
        col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod1)$r.squared, 3)),
                       paste("Adjusted Coefficient of Determination $R^2$=", round(summary(mod1)$adj.r.squared, 3)),
                       paste("Standard Error $\\\\sigma$=", round(summary(mod1)$sigma, 3))),
           escape = FALSE)
```

@@

```{r create-prediction-grid, include=FALSE}
new.data = expand.grid(parks = seq(0.36, 1.798, by = 0.01),
                       residential = seq(0.9736, 1.26, by = 0.01),
                       day = seq(37, 96, 1),
                       NY = c(0, 1)) %>%
  mutate(parks2 = parks^2,
         residential2 = residential^2,
         parksXresidential = parks * residential,
         day2 = day^2,
         parksXday = parks * day,
         parksXday2 = parks2 * day2,
         residentialXday = residential * day,
         residentialXday2 = residential * day2,
         NYxday = NY * day)

@@

```{r check-mobility-indicators-on-2020-03-21, include=FALSE}
c19 %>% filter(date == "2020-03-21") %>% select(day, parks, residential) %>% summary()
```

```{r plot-03-21, include=FALSE, fig.cap="\\label{fig:plot-03-21}Prediction surface at various levels of the mobility indicators on March 21, 2020"}
plot_03_21 <- ggplot(data = new.data %>% filter(day == 60, parks >= 0.8527 & parks <= 1.5555, residential >= 1.025 & residential <= 1.09),
       aes(x = parks,
           y = residential)) +
  geom_tile(aes(fill = log_new_cases)) + 
  geom_contour(aes(z = log_new_cases), color = "white") +
  #geom_text_contour(aes(z = incidence), stroke = 0.1, size = 3.5) +
  geom_point(data = c19 %>% filter(day == 60, state != "New York" & 
                                     state != "District of Columbia") %>% 
               select(parks, residential)) +
  geom_text(data = c19 %>% 
              filter(day == 60,
                     state == "Arkansas" |
                       state == "Mississippi" |
                       state == "Oklahoma" |
                       state == "Idaho" |
                       state == "New Jersey" |
                       state == "Michigan" |
                       state == "California") %>%
              select(parks, residential, state), 
            aes(label = state),
            nudge_y = -0.01) +
  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1) +
  theme_light() +
  ggtitle("March 21, 2020")
```

```{r check-mobility-indicators-on-2020-04-05, include=FALSE}
c19 %>% filter(date == "2020-04-05") %>% select(day, parks, residential) %>% summary()
```

```{r plot-04-05, include=FALSE, fig.cap="\\label{fig:plot-04-05}Prediction surface at various levels of the mobility indicators on April 05, 2020"}
plot_04_05 <- ggplot(data = new.data %>% filter(day == 75, parks >= 0.4545 & parks <= 1.6009, residential >= 1.122 & residential <= 1.253),
       aes(x = parks,
           y = residential)) +
  geom_tile(aes(fill = log_new_cases)) + 
  geom_contour(aes(z = log_new_cases), color = "white") +
  #geom_text_contour(aes(z = incidence), stroke = 0.1, size = 3.5) +
  geom_point(data = c19 %>% filter(day == 75, state != "New York" & 
                                     state != "District of Columbia") %>% 
               select(parks, residential)) +
  geom_text(data = c19 %>% 
              filter(day == 75,
                     state == "Arkansas" |
                       state == "Mississippi" |
                       state == "Oklahoma" |
                       state == "Idaho" |
                       state == "New Jersey" |
                       state == "Michigan" |
                       state == "California") %>%
              select(parks, residential, state), 
            aes(label = state),
            nudge_y = -0.01) +
  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1) +
  theme_light() +
  ggtitle("April 05, 2020")
```

```{r check-mobility-indicators-on-2020-04-20, include=FALSE}
c19 %>% filter(date == "2020-04-20") %>% select(day, parks, residential) %>% summary()
```

>>>>

---
title: Using Google Community Mobility Reports to investigate the growth of COVID-19 in the United States
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
address:
  - code: McMaster University
    address: School of Geography and Earth Sciences, McMaster University, Hamilton, ON, L8S 4K1, Canada
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
   In 2020 Google released a set of Community Mobility Reports (GCMR). These reports are based on the company's location-tracking capabilities and measure changes in mobility with respect to a baseline. This novel source of data offers an opportunity to investigate potential correlations between mobility and transmission of COVID-19. Using data from the New York Times on COVID-19 cases and GCMR, this paper presents an analysis of mobility levels and new daily reported cases of COVID-19 by state in the US. The results provide insights about the utility and interpretability of GCMR for COVID-19 research and decision-making.

journal: "Transportation Findings"
date: "`r Sys.Date()`"
bibliography: References.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
always_allow_html: true
---

@@

```{r add-day-variable, include=FALSE}
#add day variable:
c19 <- c19 %>%
  mutate(day = yday(date) - yday("2020-04-05"))
```

@@

```{r prepare-variables-for-analysis, include=FALSE}
#replace mobility indicators by their lagged moving averages
c19 <- c19 %>% 
  mutate(retail = Mean_r_and_r_lag11,
         groceries = Mean_g_and_p_lag11,
         parks = Mean_parks_lag11,
         parks2 = Mean_parks_lag11^2,
         transit = Mean_transit_lag11,
         work = Mean_work_lag11,
         work2 = work^2,
         residential = Mean_residential_lag11,
         residential2 = Mean_residential_lag11^2,
         parksXresidential = parks * residential,
         parksXwork = parks * work,
         day2 = day^2,
         parksXday = parks * day,
         parksXday2 = parks * day2,
         workXday = work * day,
         workXday2 = work * day2,
         residentialXday = residential * day,
         residentialXday2 = residential * day2,
         NY = ifelse(state == "New York", 1, 0),
         NYxday = NY * day)
```

@@

The main policy tool to control the spread of the COVID-19 pandemic has been stay-at-home orders, which in the United States have been implemented on a state-by-state basis, with considerable variations in compliance. Concurrently, numerous initiatives have been developed to track the progress and the impact of the pandemic. As a result, there are new sources of data such as the recently-released Google Community Mobility Reports (GCMR)\footnote{https://www.google.com/covid19/mobility/}, as well as The New York Times repository of COVID-19 data\footnote{https://github.com/nytimes/covid-19-data}. These two open data sets offer novel opportunities to investigate in quasi-real time the relationship between mobility patterns and transmission of COVID-19.

@@

This paper is a reproducible research document. The source is an `R` markdown file available in a public repository\footnote{See folder Covid-19-Google-CMR-US in \url{https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US/}}.

@@

GCMR use aggregated and anonymized data to chart changes in mobility with respect to different classes of places (see Table \ref{tab:descriptive-statistics}). Mobility indicators are calculated based on the frequency and length of visits. The reports give percentage change from a baseline level, which corresponds to the median value of mobility of identical days of the week during the period between January 3 and Feb 6, 2020. Covid-19 data is compiled by The New York Times based on reports from state and local health agencies.

```{r descriptive-statistics, echo=FALSE}
data.frame(Variable = c("New Cases",
                        "date",
                        "retail",
                        "groceries",
                        "parks",
                        "transit",
                        "work",
                        "residential"),
           Definition = c("New Daily Cases of COVID-19",
                        "Date",
                        "Mobility trends for places like restaurants, cafes, shopping centers, theme parks, museums, libraries, and movie theaters",
                        "Mobility trends for places like grocery markets, food warehouses, farmers markets, specialty food shops, drug stores, and pharmacies",
                        "Mobility trends for places like local parks, national parks, public beaches, marinas, dog parks, plazas, and public gardens",
                        "Mobility trends for places like public transport hubs such as subway, bus, and train stations",
                        "Mobility trends for places of work",
                        "Mobility trends for places of residence"),
           min = c(as.character(min(c19$new_cases)),
                   as.character(min(c19$date)),
                   as.character(round(min(c19$retail), 2)),
                   as.character(round(min(c19$groceries), 2)),
                   as.character(round(min(c19$parks), 2)),
                   as.character(round(min(c19$transit), 2)),
                   as.character(round(min(c19$work), 2)),
                   as.character(round(min(c19$residential), 2))),
           median = c(as.character(median(c19$new_cases)),
                   as.character(median(c19$date)),
                   as.character(round(median(c19$retail), 2)),
                   as.character(round(median(c19$groceries), 2)),
                   as.character(round(median(c19$parks), 2)),
                   as.character(round(median(c19$transit), 2)),
                   as.character(round(median(c19$work), 2)),
                   as.character(round(median(c19$residential), 2))),
           max = c(as.character(max(c19$new_cases)),
                   as.character(max(c19$date)),
                   as.character(round(max(c19$retail), 2)),
                   as.character(round(max(c19$groceries), 2)),
                   as.character(round(max(c19$parks), 2)),
                   as.character(round(max(c19$transit), 2)),
                   as.character(round(max(c19$work), 2)),
                   as.character(round(max(c19$residential), 2))),
           sd = c(as.character(round(sd(c19$new_cases), 2)),
                   as.character(""),
                   as.character(round(sd(c19$retail), 2)),
                   as.character(round(sd(c19$groceries), 2)),
                   as.character(round(sd(c19$parks), 2)),
                   as.character(round(sd(c19$transit), 2)),
                   as.character(round(sd(c19$work), 2)),
                   as.character(round(sd(c19$residential), 2)))) %>%
  kable("latex",
        booktabs = TRUE,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics of the data set") %>%
  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"), font_size = 7) %>%
  column_spec(2, width = "12em") %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

For analysis, all mobility indicators are centered so that the value of 1 is the base mobility, and a 0.01 deviation corresponds to a 1% change. The incubation time of the disease is between 2 and 12 days [95% interval; see @Lauer2020incubation]. Given this, it is to be expected that any changes in mobility will have a lagged effect (if any) on the discovery of new cases. For this reason, lagged moving averages of the mobility indicators are calculated. Furthermore, it is possible that mobility and reports of new cases of COVID-19 are endogenous, if the public adjust their mobility according to reports of the incidence. Therefore, in addition to being consistent with an incubation period, use of lagged indicators also helps to break this potential endogeneity.

The lagged indicators are calculated as the mean of the mobility indicator using the values from date-minus-12-days to date-minus-2-days. Furthermore, using the cumulative number of reported COVID-19 cases, the total number of new daily cases is calculated. This variable (log-transformed after adding a small constant) is paired with the corresponding lagged moving average of the mobility indicators. The log-transformation is useful to avoid negative values of daily new cases when making predictions. Table \ref{tab:descriptive-statistics} shows the descriptive statistics of the data set. Analysis is based on correlation analysis, multivariate regression, and data visualization.

@@

Table \ref{tab:correlation-analysis} shows that the mobility indicators are highly correlated with each other. Two variables are selected for multivariate analysis: parks- and work-related mobility. Work has a high correlation with the outcome variable, and its correlation with parks is relatively weak, which increases the information content of the two variables. Furthermore, parks- and work-related mobility represent two dimensions of out-of-home activities: mandatory and discretionary travel.

```{r check-correlations, echo=FALSE}
#Check the correlation between the mobility indicators and incidence:
c19 %>% 
  select(log_new_cases, retail, groceries, parks, transit, work, residential) %>%
  cor(use = "pairwise.complete.obs") %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators")  %>%
  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"), font_size = 7) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

A regression model is estimated with the log of new daily cases as the dependent variable. The covariates enter the regression in the form of a second order polynomial expansion. In addition, the date (centered on April 5) is introduced to account for the temporal trend of the pandemic. Finally, an indicator variable for the state of New York and an interaction with the date are used to distinguish the unusually high incidence of the disease there. The results appear in Table \ref{tab:model-results}. The model provides a good fit to the data and all variables reported are significant at $p<0.05$ or better.

There is an overall temporal trend that indicates a growing number of cases over time, at an accelerating rate (see positive sign of date^2). Mobility related to parks and to work both tend to increase the number of new cases; the negative sign for their interaction is indicative of the trade-offs between these two forms of mobility and their impact on new cases. The influence of parks-related mobility was relatively weaker early in the pandemic (negative sign of the parks x date term) but has become more important over time (positive sign of the parks x date^2 term). The opposite happens with work-related mobility, which started with a greater effect (positive sign of work x date term), but whose impact has declined over time (negative sign of work x date^2 term) New York has on average more new daily cases than the rest of the states, but this has declined over time (see negative sign of NY x date).

```{r model-estimation, include=FALSE}
#This is the regression model (forcing the regression through the intercept):
mod1 <- lm(log_new_cases ~  0 +
             day +
             day2 +
             parks2 + 
             parks + 
             parksXwork +
             work +
             work2 +
             parksXday + 
             parksXday2 +
             workXday +
             workXday2 +
             NY + 
             NYxday, 
           data = c19 #%>% filter(date >= "2020-04-01")
           )
summ(mod1)
```

```{r model-results, echo = FALSE}
data.frame(Variable = c("date",
                        "date^2",
                        "parks^2",
                        "parks",
                        "parks x work",
                        "work",
                        "work^2",
                        "parks x date",
                        "parks x date^2",
                        "work x date",
                        "work x date^2",
                        "NY",
                        "NY x date"),
           Coefficient = as.numeric(coefficients(mod1)),
           pval = as.numeric(summary(mod1)$coefficients[,4])) %>%
  mutate(pval = ifelse(pval >= 0.001, round(pval, 4), "<0.001")) %>%
  kable("latex",
        booktabs = TRUE,
        digits = 4,
        caption = "\\label{tab:model-results}Results of estimating regression model. Dependent variable is log(New Cases + 0.0001).",
        align = c("l", "c", "c"),
        col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7) %>%
  footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod1)$r.squared, 3)),
                       paste("Adjusted Coefficient of Determination $R^2$=", round(summary(mod1)$adj.r.squared, 3)),
                       paste("Standard Error $\\\\sigma$=", round(summary(mod1)$sigma, 3))),
           escape = FALSE)
```

Visualization is the most effective way to understand the trend according to the mobility indicators and date. Figure \ref{fig:prediction-plots} shows the prediction surfaces on three different dates: March 21, when the first states began implementing stay-at-home orders; April 5, fifteen days later; and April 20, fifteen more days later, and at a time when some states started to consider relaxing stay-at-home orders.

On March 21 there were still only minor departures from baseline mobility (recall that these are temporally lagged); the prediction surface at this point is relatively flat. This changes by April 5, when work-based mobility has declined substantially. Although every state registers lower work-based mobility, there are large variations in parks-based mobility, with some states seeing increases of up to 60% for this class of mobility.

The prediction surface indicates an expectation of a greater number of new daily cases as either class of mobility increases, but the effect is not linear. On the last date examined, April 20, the trend becomes more steep for park-based mobility, even as this indicator continues to display large variations from the baseline in both directions. The white dashed lines in the plots are the folds of the saddles, and represent, for each date, the combination of parks- and work-related mobility levels that tended to minimize the emergence of new cases.

@@

```{r create-prediction-grid, include=FALSE}
new.data = expand.grid(parks = seq(0.36, 1.9, by = 0.01),
                       work = seq(0.33, 1.08, by = 0.01),
                       day = seq(-15, 15, 1),
                       NY = c(0, 1)) %>%
  mutate(parks2 = parks^2,
         work2 = work^2,
         parksXwork = parks * work,
         day2 = day^2,
         parksXday = parks * day,
         parksXday2 = parks2 * day2,
         workXday = work * day,
         workXday2 = work * day2,
         NYxday = NY * day)

@@

```{r check-mobility-indicators-on-2020-03-21, include=FALSE}
c19 %>% filter(date == "2020-03-21") %>% select(day, parks, work) %>% summary()
```

```{r plot-03-21, include=FALSE, fig.cap="\\label{fig:plot-03-21}Prediction surface at various levels of the mobility indicators on March 21, 2020"}
plot_03_21 <- ggplot(data = new.data %>% filter(day == -15, parks >= 0.8527 & parks <= 1.5555, work >= 0.7627 & work <= 0.9409),
       aes(x = parks,
           y = work)) +
  geom_tile(aes(fill = log_new_cases)) + 
  geom_contour(aes(z = log_new_cases), color = "black", size = 0.1, binwidth = 0.1, alpha = 1) +
  geom_abline(intercept = 0.43, slope = 0.45, color = "white", linetype = "dashed", size = 0.5) +
  geom_point(data = c19 %>% filter(day == -15, state != "New York") %>% 
               select(parks, work)) +
  geom_text(data = c19 %>% 
              filter(day == -15,
                     state == "Arkansas" |
                       state == "Mississippi" |
                       state == "Oklahoma" |
                       state == "Idaho" |
                       state == "New Jersey" |
                       state == "Michigan" |
                       state == "California") %>%
              select(parks, work, state), 
            aes(label = state),
            nudge_y = -0.01) +
#  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1, limits = c(4,15)) +
  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1) +
  theme_light() +
  ggtitle("March 21, 2020")
```

```{r check-mobility-indicators-on-2020-04-05, include=FALSE}
c19 %>% filter(date == "2020-04-05") %>% select(day, parks, work) %>% summary()
```

```{r plot-04-05, include=FALSE, fig.cap="\\label{fig:plot-04-05}Prediction surface at various levels of the mobility indicators on April 05, 2020"}
plot_04_05 <- ggplot(data = new.data %>% filter(day == 0, parks >= 0.4545 & parks <= 1.6009, work >= 0.3673 & work <= 0.6955),
       aes(x = parks,
           y = work)) +
  geom_tile(aes(fill = log_new_cases)) + 
  geom_contour(aes(z = log_new_cases), color = "black", size = 0.1, binwidth = 0.1, alpha = 1) +
  geom_abline(intercept = 0.162, slope = 0.45, color = "white", linetype = "dashed", size = 0.5) +
  geom_point(data = c19 %>% filter(day == 0, state != "New York") %>% 
               select(parks, work)) +
  geom_text(data = c19 %>% 
              filter(day == 0,
                     state == "Arkansas" |
                       state == "Mississippi" |
                       state == "Oklahoma" |
                       state == "Idaho" |
                       state == "New Jersey" |
                       state == "Michigan" |
                       state == "California") %>%
              select(parks, work, state), 
            aes(label = state),
            nudge_y = -0.01) +
  #scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1, limits = c(4,15)) +
  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1) +
  theme_light() +
  ggtitle("April 05, 2020")
```

```{r check-mobility-indicators-on-2020-04-20, include=FALSE}
c19 %>% filter(date == "2020-04-20") %>% select(day, parks, work) %>% summary()
```
